DeepSphere: a graph-based spherical CNN

### 2019-09-25 main.tex
intro ready for submission

\todo{
\begin{itemize}
	\item shall we really name it? Or simply reference the general idea of a graph-based spherical CNN.
	\item adjacency matrix: $A$ instead of $W$?
	\item anonymize DeepSphere citation
% 	\item climate spherical figure: inputs with TC and AR labels superimposed [Frédérick]
	\item citet for inline, citep for parenthesis
	\item More details on related work? Not sure we have the space. If yes, detailed comparison with related work after method.
	\item PSD of climate data: how does it compare to cosmo and 3D objects?
	\item align the spheres in figure 1
	\item mention that \citep{esteves2018sphericalcnn} and \citep{cohen2018sphericalcnn} computed equivariance error
	\item check footnotes
	\item climate data on spectrum plot
	\item result tables: only what supports the main message in the text, everything we have in the appendix
	\item Unify vocabulary. Sampling, discretization, or pixelization of the sphere? Signal, maps, functions. Irregular, uneven. Coarsening, down-sampling.
	\item check English vs American English
	\item unify enumerations. I like (i), (ii), etc.
	\item notation: should we use indices from 1 to n instead of 0 to n-1 for compactness?
	\item add \citet{bruna2013gnn} in equivariance? They were the first to do a spherical CNN with graphs. Graph is built from a uniform random sampling of the sphere.
	\item figure placements
	\item do the experiments on the icosahedral grid with the pooling as defined in the paper
	\item put all the code (learning experiments from Frédérick and equivariance experiments from Martino) in a common reproducible git repository to be linked from the paper
	\item update pygsp and merge sphere branch
	\item check that NN architectures and hyper-params are correctly reported
	\item keep NN architecture and hyper-params in the main text? Move the paragraph to appendix if not enough space.
	%\item shall we still site adam and batch norm? no, mainstream enough
	\item check results for ModelNet40 and integrate in text
	\item Check that HEALPix is introduced in the introduction
	\item heuristic used to set the kernel width in v1. Add values to \figref{kernel_widths} for comparison.
	\item check experiments section [Frédérick]: all the results numbers and NN + training details
	\item compute the invariance error of the whole NNs (before and after training, invariance can be learned)
	\item add it with equivariance error of the convolution to \tabref{cosmo}
\end{itemize}
}
\newpage

\begin{abstract}
Selecting the right convolution for a spherical neural network is a delicate tradeoff between its efficiency and its exactictude with respect of equivariance. DeepSphere, as a graph based method strikes as a controllable balance between these two aspects. This contribution 1) studies empirically and theoretically how the construction of its underlying graph affects equivariance, and 2) evaluates its performance on a large set of relevant problems.
Comparison with previous work on shape classification and climate event detection suggests that anisotropic filters are an unnecessary price to pay.

@@

Experiments on xxx show
Comparison with previous work on shape classification and climate event detection suggests that anisotropic filters are an unnecessary price to pay.
% memory, computations, implementation complexity

\end{abstract}

@@

Spherical data is found in many applications.
% spherical data f: S^2 -> R, projection from R^3
Planetary data (such as meteorological or geological measurements) and brain activity are example of intrinsically spherical data.
The observation of the universe, LIDAR scans, and the digitalization of 3D objects are examples of projections due to observation.
See \figref{examples}.
In many applications, labels or variables are to be inferred from spherical data.
Examples are the inference of cosmological parameters from the distribution of mass in the universe \citep{perraudin2019deepspherecosmo}, the segmentation of omnidirectional images \citep{khasanova2017sphericalcnn}, and the segmentation of cyclones from Earth observation \citep{mudigonda2017climateevents}.

\begin{figure}[h]
	\centering
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_brain_meg}
	} \hspace{0.1em}
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_cosmo_cmb}
	} \hfill
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_climate_TMQ}
		\label{fig:examples:climate}
	} \hfill
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_ghcn_daily_tmax}
		\label{fig:examples:ghcn:tmax}
	} \hfill
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_ghcn_graph}
		\label{fig:examples:ghcn:graph}
	}
	\caption{
		Examples of intrinsically spherical data:
		(a) brain activity recorded through magnetoencephalography (MEG),\protect\footnotemark
		(b) the cosmic microwave background (CMB) temperature from \citet{planck2015overview},
		(c) hourly precipitation from a climate simulation \citep{jiang2019sphericalcnn}, %\citep{mudigonda2017climateevents},
		(d) daily maximum temperature from the Global Historical Climatology Network (GHCN).\protect\footnotemark
		A rigid full-sphere pixelization is not ideal: brain activity is only measured on the scalp, the Milky Way's galactic plane masks observations, climate scientists desire a variable resolution, and the position of weather stations is arbitrary and changes over time.
		(e) Graphs can faithfully and efficiently represent sampled spherical data by placing vertices where data has been measured.
		%\todo{something from climate, weather?}
		%\todo{compare ideal full-spheres (like SHREC'17 projection from Esteves, 360 from Coors) to real-world measurements?}
		%\todo{horizontal colorbar for brain}
	}
	\label{fig:examples}
\end{figure}
\footnotetext[1]{\scriptsize\url{https://martinos.org/mne/stable/auto_tutorials/plot_visualize_evoked.html}}
\footnotetext[2]{\scriptsize\url{https://www.ncdc.noaa.gov/ghcn-daily-description}}

@@

% keep the story simple and focus on the tradeoff between cost (time / computational and storage / memory) and exactitude
As neural networks (NNs) have proved to be great tools for inference, variants have been developed to handle spherical data.
Exploiting the locally Euclidean property of the sphere, early attempts used standard 2D convolutions on a grid discretization of the sphere \citep{boomsma2017sphericalcnn, su2017sphericalcnn, coors2018sphericalcnn}.
While simple and efficient, those convolutions are not equivariant to rotations.
On the other side of this tradeoff, \citet{cohen2018sphericalcnn} and \citet{esteves2018sphericalcnn} proposed to perform proper spherical convolutions through the spherical harmonic transform.
While equivariant to rotations, those convolutions are expensive.

@@

As the lack of equivariance penalizes performance (\secref{exp:cosmo}) and expensive convolutions prohibit their application to some real-world problems, methods standing between those two extremes are desired.
\citet{cohen2019gauge} proposed to reduce costs by limiting the size of the representation of the symmetry group.
% (from the continuous rotation group SO(3) to a discrete rotation group)
% the dimensionality of feature maps is equal to the size of the irreducible representation of the symmetry group
%As a platonic solid, it \todo{exhibits discrete rotations}.
% the representation of its symmetry group is much smaller than SO(3)
\todo{not all rotations, subset of SO(3)?}
The projection of the data from the sphere to the icosahedron however introduces distortions that might hinder performance (see the experiments in \secref{exp:climate}).
\todo{Where does Jiang fit? Need a global coordinate system (ok for planets, not projections like cosmo)}

Another approach represents the discretized sphere as a graph connecting pixels according to the distance between them \citep{bruna2013gnn, khasanova2017sphericalcnn, perraudin2019deepspherecosmo}.
While Laplacian-based graph convolutions are more efficient than spherical convolutions, they are not exactly equivariant \citep{defferrard2019deepsphereequiv}.
In this work, we argue that graph-based spherical CNNs strike an interesting balance, with a controllable tradeoff between cost and equivariance.
Experiments on multiple problems of practical interest show the competitiveness and flexibility of this approach.
% Furthermore, experiments show that, perhaps surprisingly and contrarily to previously published results \citep{cohen2019gauge} \todo{[Bronstein?]}, anisotropic filters might be an unnecessary price to pay.
% This paper focuses on improving the rotation equivariance of \citet{perraudin2019deepspherecosmo}, a scalable graph-based method designed for cosmological applications.

@@

\paragraph{Convolution.}
On Euclidean domains, convolutions are efficiently implemented by sliding a window in the signal domain.
On the sphere however, there is no straightforward way to implement a convolution in the signal domain due to non-uniform samplings.
Convolutions are most often performed in the spectral domain through a spherical harmonic transform (SHT).
That is the approach taken by \citet{cohen2018sphericalcnn} and \citet{esteves2018sphericalcnn}, which has a computational cost of $\bO(n^{3/2})$ on isolatitude pixelizations (such as the HEALPix and equiangular grids) and $\bO(n^2)$ in general.
On the other hand, following \citet{defferrard2016graphnn}, graph convolutions can be defined as
\begin{equation} \label{eqn:graph_conv}
	h(\b{L}) \b{f} = \left(\sum_{i=0}^K \alpha_i \b{L}^i\right) \b{f},
\end{equation}
where $K$ is the polynomial order (which corresponds to the filter's size) and $\alpha_i$ are the coefficients to be optimized during training.\footnote{In practice, training with Chebyshev polynomials (instead of monomials) is slightly more stable. We believe it to be due to their orthogonality and uniformity.}
Those convolutions are used by \citet{khasanova2017sphericalcnn} and \citet{perraudin2019deepspherecosmo} and cost $\bO(n)$ operations through a recursive application of $\b{L}$.\footnote{As long as the graph is sparsified such that the number of edges, i.e., the number of non-zeros in $\b{A}$, is proportional to the number of vertices $n$. This can always be done as most weights are very small.}

>>>>

@@

\todo{
\begin{itemize}
	\item shall we really name it? Or simply reference the general idea of a graph-based spherical CNN.
	\item adjacency matrix: $A$ instead of $W$?
	\item anonymize DeepSphere citation
% 	\item climate spherical figure: inputs with TC and AR labels superimposed [Frédérick]
	\item citet for inline, citep for parenthesis
	\item More details on related work? Not sure we have the space. If yes, detailed comparison with related work after method.
	\item PSD of climate data: how does it compare to cosmo and 3D objects?
	\item align the spheres in figure 1
	\item mention that \citep{esteves2018sphericalcnn} and \citep{cohen2018sphericalcnn} computed equivariance error
	\item check footnotes
	\item climate data on spectrum plot
	\item result tables: only what supports the main message in the text, everything we have in the appendix
	\item Unify vocabulary. Sampling, discretization, or pixelization of the sphere? Signal, maps, functions. Irregular, uneven. Coarsening, down-sampling.
	\item check English vs American English
	\item unify enumerations. I like (i), (ii), etc.
	\item notation: should we use indices from 1 to n instead of 0 to n-1 for compactness?
	\item add \citet{bruna2013gnn} in equivariance? They were the first to do a spherical CNN with graphs. Graph is built from a uniform random sampling of the sphere.
	\item figure placements
	\item do the experiments on the icosahedral grid with the pooling as defined in the paper
	\item put all the code (learning experiments from Frédérick and equivariance experiments from Martino) in a common reproducible git repository to be linked from the paper
	\item update pygsp and merge sphere branch
	\item check that NN architectures and hyper-params are correctly reported
	\item keep NN architecture and hyper-params in the main text? Move the paragraph to appendix if not enough space.
	%\item shall we still site adam and batch norm? no, mainstream enough
	\item check results for ModelNet40 and integrate in text
	\item Check that HEALPix is introduced in the introduction
	\item heuristic used to set the kernel width in v1. Add values to \figref{kernel_widths} for comparison.
	\item check experiments section [Frédérick]: all the results numbers and NN + training details
	\item compute the invariance error of the whole NNs (before and after training, invariance can be learned)
	\item add it with equivariance error of the convolution to \tabref{cosmo}
	\item insert \citet{jiang2019sphericalcnn} in intro: Need a global coordinate system (ok for planets, not projections like cosmo)
\end{itemize}
}
\newpage

\begin{abstract}
Selecting the right convolution for a spherical neural network is a delicate tradeoff between its efficiency and its exactictude with respect of equivariance. DeepSphere, as a graph based method strikes as a controllable balance between these two aspects. This contribution 1) studies empirically and theoretically how the construction of its underlying graph affects equivariance, and 2) evaluates its performance on a large set of relevant problems.
Comparison with previous work on shape classification and climate event detection suggests that anisotropic filters are an unnecessary price to pay.
% memory, computations, implementation complexity
\end{abstract}

@@

Spherical data is found in many applications (\figref{examples}).
% spherical data f: S^2 -> R, projection from R^3
Planetary data (such as meteorological or geological measurements) and brain activity are example of intrinsically spherical data.
The observation of the universe, LIDAR scans, and the digitalization of 3D objects are examples of projections due to observation.
Labels or variables are often to be inferred from them.
Examples are the inference of cosmological parameters from the distribution of mass in the universe \citep{perraudin2019deepspherecosmo}, the segmentation of omnidirectional images \citep{khasanova2017sphericalcnn}, and the segmentation of cyclones from Earth observation \citep{mudigonda2017climateevents}.

\begin{figure}[h]
	\centering
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_brain_meg}
	} \hspace{0.1em}
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_cosmo_cmb}
	} \hfill
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_climate_TMQ}
		\label{fig:examples:climate}
	} \hfill
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_ghcn_daily_tmax}
		\label{fig:examples:ghcn:tmax}
	} \hfill
	\subfloat[]{
		\includegraphics[height=0.20\linewidth,valign=t]{example_ghcn_graph}
		\label{fig:examples:ghcn:graph}
	}
	\caption{
		Examples of spherical data:
		(a) brain activity recorded through magnetoencephalography (MEG),\protect\footnotemark
		(b) the cosmic microwave background (CMB) temperature from \citet{planck2015overview},
		(c) hourly precipitation from a climate simulation \citep{jiang2019sphericalcnn}, %\citep{mudigonda2017climateevents},
		(d) daily maximum temperature from the Global Historical Climatology Network (GHCN).\protect\footnotemark
		A rigid full-sphere pixelization is not ideal: brain activity is only measured on the scalp, the Milky Way's galactic plane masks observations, climate scientists desire a variable resolution, and the position of weather stations is arbitrary and changes over time.
		(e) Graphs can faithfully and efficiently represent sampled spherical data by placing vertices where it matters.
		%\todo{horizontal colorbar for brain}
	}
	\label{fig:examples}
\end{figure}
\footnotetext[1]{\scriptsize\url{https://martinos.org/mne/stable/auto_tutorials/plot_visualize_evoked.html}}
\footnotetext[2]{\scriptsize\url{https://www.ncdc.noaa.gov/ghcn-daily-description}}

@@

% keep the story simple and focus on the tradeoff between cost (time / computational and storage / memory) and exactitude
As neural networks (NNs) have proved to be great tools for inference, variants have been developed to handle spherical data.
Exploiting the locally Euclidean property of the sphere, early attempts used standard 2D convolutions on a grid discretization of the sphere \citep{boomsma2017sphericalcnn, su2017sphericalcnn, coors2018sphericalcnn}.
While simple and efficient, those convolutions are not equivariant to rotations.
On the other side of this tradeoff, \citet{cohen2018sphericalcnn} and \citet{esteves2018sphericalcnn} proposed to perform proper spherical convolutions through the spherical harmonic transform.
While equivariant to rotations, those convolutions are expensive (\secref{method:convolution}).

@@

As a lack of equivariance can penalize performance (\secref{exp:cosmo}) and expensive convolutions prohibit their application to some real-world problems, methods standing between those two extremes are desired.
\citet{cohen2019gauge} proposed to reduce costs by limiting the size of the representation of the symmetry group by projecting the data from the sphere to the icosahedron.
The distortions introduced by this projection might however hinder performance (\secref{exp:climate}).
% (from the continuous rotation group SO(3) to a discrete rotation group)
% the dimensionality of feature maps is equal to the size of the irreducible representation of the symmetry group
%As a platonic solid, it \todo{exhibits discrete rotations}.
% the representation of its symmetry group is much smaller than SO(3)
%\todo{not all rotations, subset of SO(3)?}

Another approach is to represent the discretized sphere as a graph connecting pixels according to the distance between them \citep{bruna2013gnn, khasanova2017sphericalcnn, perraudin2019deepspherecosmo}.
While Laplacian-based graph convolutions are more efficient than spherical convolutions, they are not exactly equivariant \citep{defferrard2019deepsphereequiv}.
In this work, we argue that graph-based spherical CNNs strike an interesting balance, with a controllable tradeoff between cost and performance (linked to equivariance).
Experiments on multiple problems of practical interest show the competitiveness and flexibility of this approach.
% Furthermore, experiments show that, perhaps surprisingly and contrarily to previously published results \citep{cohen2019gauge} \todo{[Bronstein?]}, anisotropic filters might be an unnecessary price to pay.
% This paper focuses on improving the rotation equivariance of \citet{perraudin2019deepspherecosmo}, a scalable graph-based method designed for cosmological applications.

@@

\paragraph{Convolution.} \label{sec:method:convolution}
On Euclidean domains, convolutions are efficiently implemented by sliding a window in the signal domain.
On the sphere however, there is no straightforward way to implement a convolution in the signal domain due to non-uniform samplings.
Convolutions are most often performed in the spectral domain through a spherical harmonic transform (SHT).
That is the approach taken by \citet{cohen2018sphericalcnn} and \citet{esteves2018sphericalcnn}, which has a computational cost of $\bO(n^{3/2})$ on isolatitude pixelizations (such as the HEALPix and equiangular grids) and $\bO(n^2)$ in general.
On the other hand, following \citet{defferrard2016graphnn}, graph convolutions can be defined as
\begin{equation} \label{eqn:graph_conv}
	h(\b{L}) \b{f} = \left(\sum_{i=0}^K \alpha_i \b{L}^i\right) \b{f},
\end{equation}
where $K$ is the polynomial order (which corresponds to the filter's size) and $\alpha_i$ are the coefficients to be optimized during training.\footnote{In practice, training with Chebyshev polynomials (instead of monomials) is slightly more stable. We believe it to be due to their orthogonality and uniformity.}
Those convolutions are used by \citet{khasanova2017sphericalcnn} and \citet{perraudin2019deepspherecosmo} and cost $\bO(n)$ operations through a recursive application of $\b{L}$.\footnote{As long as the graph is sparsified such that the number of edges, i.e., the number of non-zeros in $\b{A}$, is proportional to the number of vertices $n$. This can always be done as most weights are very small.}
